Microclimatic spatial variation buffers local warming in alpine vegetation

### 2022-01-20 doc/manuscript.Rmd
species changes

The study was conducted in the calcareous massif of Picos de Europa National Park, in northern Spain (Figure 1A). The study area is a biodiversity hotspot for cold-adapted plants in the Iberian Peninsula and a biogeographical hub for Alpine and Mediterranean lineages in Western Europe (XXX). The main calcareous massif occupies c. 50 km^2^ and support high diversity of ecosystems, with alpine vegetation mostly occurring between 1900 and 2400 m a.s.l., with a local species pool of XX species (Jiménez-Alfaro et al. XXX). In 2008, we established a long-term ecological research program for monitoring soil climate and vegetation change. We selected four study sites along a north-south gradient, reflecting the main macroclimatic variation from oceanic to Mediterranean influence. The sites were selected to... general design and description of sites... (Borja)

@@

In the temporal survey (2 visits x 2 plots per site, n = 16) we recorded 42 species in 2009 and 45 in 2019. Of the species recorded in 2009, we did not find again in 2019 the following three: *Festuca burnatii*, *Galium pyrenaicum* and *Viola riviniana*. Conversely, in 2019 we recorded six species that we had not seen in 2009: *Arenaria purpurascens*, *Gentiana verna*, *Potentilla crantzii*, *Sedum brevifolium* and *Seseli montanum*. Changes in species frequency were/were not correlated with XX and XX.... (Table 1) (Eduardo). In the spatial survey (20 plots per site, n = 80; including one plot placed in rocks with no vascular plants) we recorded 81 species, 37 of which were absent from the temporal survey plots. Across the whole study system, the five most frequent species were *Thymus praecox* subsp. *ligusticus* (83 occurences), *Anthyllis vulneraria* (73), *Koeleria vallesiana* (59), *Minuartia verna* (55) and *Helianthemum canum* (52).

>>>>

The study was conducted in the central calcareous massif of the Picos de Europa National Park, in northern Spain (Figure 1A). The study area is a biodiversity hotspot for cold-adapted plants in the Iberian Peninsula and a biogeographical hub for Alpine and Mediterranean lineages in Western Europe (XXX). The central calcareous massif occupies c. 50 km^2^ and support high diversity of ecosystems, with alpine vegetation mostly occurring between 1900 and 2400 m a.s.l., with a local species pool of XX species (Jiménez-Alfaro et al. XXX). In 2008, we established a long-term ecological research program for monitoring soil climate and vegetation change. We selected four study sites along a north-south gradient, reflecting the main macroclimatic variation from oceanic to Mediterranean influence. The sites were selected to... general design and description of sites... (Borja)

@@

Across the whole study system (temporal and spatial surveys) we recorded 86 species of vascular plants (considering *Helianthemum apenninum* subsp. *urrielense* and *Helianthemum apenninum* subsp. *cantabricum* as separate species), representing % of the total species pool of the study region. The five most frequent species were *Thymus praecox* subsp. *ligusticus* (83 occurences), *Anthyllis vulneraria* (73), *Koeleria vallesiana* (59), *Minuartia verna* (55) and *Helianthemum canum* (52). Average species richness per 1m^2^ plot was 13, with the richest plot having 25 species and the poorest one just two.

In the temporal survey (2 visits x 2 plots per site, n = 16) we recorded 42 species in 2009 and 47 in 2019. Of the species recorded in 2009, we did not find again in 2019 the following three: *Festuca burnatii*, *Galium pyrenaicum* and *Iberis carnosa*. Conversely, in 2019 we recorded eight species that we had not seen in 2009: *Arenaria purpurascens*, *Lotus corniculatus*, *Potentilla crantzii*, *Sedum album*, *Sedum brevifolium*, *Seseli montanum*, *Silene ciliata* and *Solidago virgaurea*. Of these species that appeared or disappeared from the temporal survey plots, *Arenaria purpurascens* appeared in 12 10x10 cm cells in 2019, while the rest of the cases amounted to less than 10 cells. The five species with the highest decrease in frequency from 2009 to 2019 (ignoring annual species and species that occurred in less than 10 10x10 cm cells in 2009) were *Armeria cantabrica* (85% decrease in frequency, present in 13 cells in 2009), *Poa alpina* (-83%, 18 cells), *Salix breviserrata* (-48%, 25 cells), *Jurinea humilis* (-26%, 23 cells) and *Ranunculus parnassiifolius* subsp. *favargeri* (-18%, 72 cells). The five species with the highest increases (again, ignoring annual species and species that occurred in less than 10 10x10 cm cells in 2009) were *Minuartia verna* (+278%, 19 cells), *Helianthemum apenninum* subsp. *urrielense* (+87%, 63 cells), *Arenaria moehringioides* (+85%, 13 cells), *Saxifraga conifera* (+83%, 24 cells) and *Silene acaulis* (+39%, 18 cells).

Changes in species frequency were/were not correlated with XX and XX.... (Table 1) (Eduardo). In the spatial survey (20 plots per site, n = 80; including one plot placed in rocks with no vascular plants) we recorded 81 species, 36 of which were absent from the temporal survey plots.
